---
title: "RRHO2 Analysis of WT and HuR"
author: "Sivakumar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document performs an analysis using **RRHO2** to compare gene expression in WT and HuR datasets.

## Setup Environment

```{r load-libraries}
library(readxl)
library(dplyr)
library(devtools)
library(RRHO2)
library(ggplot2)
library(grid)
```

## Load Data

```{r load-data}
wt_file <- "WT_E20_vs_WT_DMSO_deseq2.xls"
HuR_file <- "HuRKO_E20_vs_HuRKO_DMSO_deseq2.xls"

wt_data <- read_excel(wt_file, header = TRUE)
HuR_data <- read_excel(HuR_file, header = TRUE)
```

## Data Filtering

```{r filter-data}
wt_data_filtered <- wt_data %>%
  filter(!is.na(pvalue) & pvalue > 0, abs(log2FoldChange) > 1) %>%
  mutate(rank_metric = -log10(pvalue) * sign(log2FoldChange)) %>%
  arrange(desc(rank_metric))

HuR_data_filtered <- HuR_data %>%
  filter(!is.na(pvalue) & pvalue > 0, abs(log2FoldChange) > 1) %>%
  mutate(rank_metric = -log10(pvalue) * sign(log2FoldChange)) %>%
  arrange(desc(rank_metric))
```

## Find Common Genes

```{r find-common-genes}
common_genes <- intersect(wt_data_filtered$GeneID, HuR_data_filtered$GeneID)
wt_common <- wt_data_filtered %>% filter(GeneID %in% common_genes)
HuR_common <- HuR_data_filtered %>% filter(GeneID %in% common_genes)

nrow(wt_common)  # Total genes in WT dataset
nrow(HuR_common)  # Total genes in HuR dataset
length(common_genes)  # Number of common genes

wt_rank <- setNames(wt_common$rank_metric, wt_common$GeneID)
HuR_rank <- setNames(HuR_common$rank_metric, HuR_common$GeneID)

wt_rank <- wt_rank[common_genes]
HuR_rank <- HuR_rank[common_genes]

length(wt_rank)  
length(HuR_rank)  
```

## Rank Genes and Run RRHO2

```{r rrho2-analysis}
wt_rank_df <- data.frame(GeneID = names(wt_rank), rank_metric = wt_rank)
HuR_rank_df <- data.frame(GeneID = names(HuR_rank), rank_metric = HuR_rank)

RRHO_obj <- RRHO2_initialize(
  wt_rank_df, HuR_rank_df,
  labels = NULL,
  log10.ind = TRUE,
  multipleTesting = "none",
  boundary = 0.1,
  method = "hyper"
)
```

## Results Visualization

```{r rrho2-plots}
RRHO2_vennDiagram(RRHO_obj, type="uu")
RRHO2_heatmap(RRHO_obj)
```

## Summary Statistics

```{r summary-stats}
summary(RRHO_obj)
```

## Save Heatmap

```{r save-plot}
tiff("RRHO2_wt_HuR_1.png", width = 8, height = 6, units = "in", res = 100)
RRHO2_heatmap(RRHO_obj, labels = c("WT", "HuRKO"))
dev.off()
```

## Save Enhanced Heatmap

```{r save-plot2}
tiff("RRHO2_wt_HuR-1.png", width = 8, height = 6, units = "in", res = 100)

RRHO2_heatmap <- function(RRHO_obj, maximum=NULL, minimum=NULL, 
                          colorGradient=NULL, labels=NULL, 
                          tickLabelSize=2,   # increase tick font size
                          titleSize=2,         # increase title font size
                          titleMargin=0.1,       # increase margin for title
                          cbWidth = 4, 
                          ...) {
  
  hypermat <- RRHO_obj$hypermat
  method <- RRHO_obj$method
  
  if(is.null(labels)){
    labels <- RRHO_obj$labels	
  }
  
  if(!is.null(maximum)){
    hypermat[hypermat > maximum] <- maximum
  } else {
    maximum <- max(hypermat, na.rm=TRUE)
  }
  
  if(!is.null(minimum)){
    hypermat[hypermat < minimum] <- minimum
  } else {
    minimum <- min(hypermat, na.rm=TRUE)
  }
  
  if(minimum > maximum){
    stop("minimum > maximum, please check these function arguments!")
  }
  
  # Internal function to draw a colorbar
  color.bar <- function(lut, min, max=-min, 
                        nticks=11, 
                        ticks=seq(min, max, len=nticks), 
                        title='') {
    scale  <- (length(lut)-1)/(max-min)
    plot(c(0,10), c(min,max), type='n', bty='n', 
         xaxt='n', xlab='', yaxt='n', ylab='')
    mtext(title, side=4, line=titleMargin, cex=titleSize)
    axis(2, at=ticks, labels=round(ticks,0), las=1, cex.axis=tickLabelSize)
    for (i in 1:(length(lut)-1)) {
      y  <- (i-1)/scale + min
      rect(0, y, cbWidth, y+1/scale, col=lut[i], border=NA)
    }
  }
  
  if(is.null(colorGradient)){
    jet.colors  <- colorRampPalette(
      c("#00007F", "blue", "#007FFF", "cyan", 
        "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))
    colorGradient <- jet.colors(101)
  }
  
  layout(matrix(c(rep(1, 6), 2), 1, 7, byrow = TRUE))
  
  breaks <- seq(minimum, maximum, length.out = length(colorGradient) + 1)
  
  # Plot the heatmap (without default axis labels)
  image(x = 1:ncol(hypermat), y = 1:nrow(hypermat), z = t(hypermat),
        col = colorGradient, breaks = breaks, axes = FALSE,
        xlab = "", ylab = "", ...)
  
  # Suppress default mtext labels by passing empty labels
  if(!is.null(labels) && any(labels != "")){
    mtext(labels[2], side=2, line=0.5)
    mtext(labels[1], side=1, line=0.5)
  }
  
  if(method == "hyper"){
    atitle <- ifelse(RRHO_obj$log10.ind, "-log10(P-value)", "-log(P-value)")
    color.bar(colorGradient, min = minimum, max = maximum, nticks = 6, title = atitle)
  } else if (method == "fisher"){
    atitle <- "log Odds"
    color.bar(colorGradient, min = minimum, max = maximum, nticks = 6, title = atitle)
  } else {
    stop("internal error (1), please report this error to https://github.com/RRHO2/RRHO2/issues")
  }
  invisible(hypermat)
}

# Create the heatmap without default axis labels
RRHO2_heatmap(RRHO_obj, labels = c("", ""))

grid.text("WT", x = unit(0.45, "npc"), y = unit(0.03, "npc"),
          gp = gpar(fontsize = 24))

grid.text("HuRKO", x = unit(0.02, "npc"), y = unit(0.55, "npc"), rot = 90,
          gp = gpar(fontsize = 24, fontface = "italic"))

grid.text("Up", x = unit(0.04, "npc"), y = unit(0.75, "npc"),
          rot = 90, gp = gpar(fontsize = 24))
grid.text("Down", x = unit(0.04, "npc"), y = unit(0.3, "npc"),
          rot = 90, gp = gpar(fontsize = 24))
grid.text("Down", x = unit(0.25, "npc"), y = unit(0.08, "npc"),
          gp = gpar(fontsize = 24))
grid.text("Up", x = unit(0.65, "npc"), y = unit(0.08, "npc"),
          gp = gpar(fontsize = 24))

# Close graphics device
dev.off()
```

## Conclusion

This analysis provides insights into gene expression similarities and differences between WT and HuR using RRHO2.
